---
title: "Sentinel Project Analyses"
output:
  html_document: default
  pdf_document: default
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
```


## Observations

```{r}
data <-read.csv("./data/observations-7.13.26.csv")
#data <- data[c(1,4,8,14,43,47:60)]
```

```{r}
data <- data %>%
  mutate(
    location = case_when(
      str_starts(field.plant.number..from.tag.or.map., "POT") ~ "Port of Tacoma",
      TRUE ~ "WSU"
  )
) %>%
  mutate(
    field.plant.health.condition = case_when(
      is.na(field.plant.health.condition) ~ "No Selection",
      field.plant.health.condition == "" ~ "No Selection",
      TRUE ~ field.plant.health.condition
    ) 
  ) %>%
   mutate(
    scientific_name = str_trim(scientific_name),
    scientific_name = recode(
      scientific_name,
      "Quercus garryana garryana" = "Quercus garryana",
      "Pseudotsuga" = "Pseudotsuga menziesii"
    )
  ) %>%
   filter(
    !is.na(scientific_name),
    scientific_name != "",
    !is.na(`field.plant.number..from.tag.or.map.`),
    `field.plant.number..from.tag.or.map.` != ""
  ) %>% 
   mutate(
    observed_on = as.Date(observed_on),
    scientific_name = factor(scientific_name),
    field.plant.health.condition = factor(field.plant.health.condition),
    field.foliage...insect.feeding = factor(field.foliage...insect.feeding),
    field.foliage...deformation = factor(field.foliage...deformation),
    field.woody.tissue....structural.damage = factor(field.woody.tissue....structural.damage),
    field.woody.tissue...signs.of.disease = factor(field.woody.tissue...signs.of.disease),
    field.woody.tissue...insect.activity = factor(field.woody.tissue...insect.activity),
    field.woody.tissue...symptoms = factor(field.woody.tissue...symptoms),
    field.root.collar...signs.of.disease = factor(field.root.collar...signs.of.disease),
    field.critical.root.zone..within.canopy.drip.line....damage.or.concerns =
    factor(field.critical.root.zone..within.canopy.drip.line....damage.or.concerns),
    location = factor(location) 
    )%>%
   mutate(
    field.plant.number..from.tag.or.map. =
      field.plant.number..from.tag.or.map. %>%
      str_replace(regex("^wsup", ignore_case = TRUE), "WSUP") %>%  # fix capitalization
      str_replace("^WSUP\\.", "WSUP") %>%                          # remove period
      str_replace("^WSUP(\\d)", "WSUP \\1") %>%                    # add missing space
      str_squish()                                                 # clean extra spaces
      )
```



```{r}
Plant.Summary <- data %>%
  group_by(location) %>%
  summarise(
    Observations = n(),
    Unique.Plants = n_distinct(field.plant.number..from.tag.or.map.)
  )
```

**At WSU,  `r Plant.Summary$Unique.Plants[Plant.Summary$location=="WSU"]` sentinel plants were monitored through `r Plant.Summary$Observations[Plant.Summary$location=="WSU"]` observations. **

**At the Port of Tacoma, `r Plant.Summary$Unique.Plants[Plant.Summary$location=="Port of Tacoma"]` sentinel plants were monitored through `r Plant.Summary$Observations[Plant.Summary$location=="Port of Tacoma"]` observations.**

## Data Visualization


### Observations by date

```{r}
obs.by.date <- data %>%
 mutate(observed_on = as.Date(observed_on)) %>%
  count(observed_on, location)
```


```{r fig.height=5, fig.width=10}
pd <- position_dodge2(width = 0.9, preserve = "single")
ggplot(obs.by.date,
       aes(observed_on, n, fill = location)) +
  geom_col(
    position = pd,
    width = 0.8
  ) +
  geom_text(
    aes(label = n),
    position = pd,
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_bw() +
  labs(
    x = "Observation Date",
    y = "Number of Plants Monitored",
    fill = "Location"
  )
```



```{r}
most.recent.data <- data %>%
  mutate(observed_on = as.Date(observed_on)) %>%
  group_by(field.plant.number..from.tag.or.map.) %>%
  slice_max(observed_on, n = 1, with_ties = FALSE) %>%
  ungroup()
```


### Most Recent Data

The plots below filter out the first observations of each plant if it was measured more than once so far this summer. 

```{r fig.height=5, fig.width=10}
ggplot(
  most.recent.data,
  aes(field.plant.health.condition,
      fill = field.plant.health.condition)
) +
  geom_bar() +
  stat_count(
    aes(label = after_stat(count)),
    geom = "text",
    hjust = -0.2
  ) +
  coord_flip() +
  facet_wrap(~location) +
  theme_bw()+ 
  guides(fill="none")+
  labs(    
    x = "Health Condition",
    y = "Number of plants"
    )
```

```{r fig.height=5, fig.width=10}
ggplot(
  most.recent.data,
  aes(
    x = scientific_name,
    fill = field.plant.health.condition
  )
) +
  geom_bar() +
  facet_wrap(~ location,nrow=1) +
  coord_flip() +
  scale_x_discrete(labels = function(x) parse(text = paste0("italic('", x, "')"))) +
  theme_bw() +
  labs(
    x = "Species",
    y = "Number of plants",
    fill = "Health Condition"
  )
```

```{r}
Health.Conflicts <- data %>%
  group_by(`field.plant.number..from.tag.or.map.`) %>%
  filter(n_distinct(`field.plant.health.condition`, na.rm = TRUE) > 1) %>%
  arrange(`field.plant.number..from.tag.or.map.`)
```

\newpage

## Observations where condition changed

```{r}
Health.Conflict.Summary <- data %>%
  group_by(`field.plant.number..from.tag.or.map.`,scientific_name) %>%
  summarise(
    n_entries = n(),
    n_health_conditions = n_distinct(`field.plant.health.condition`, na.rm = TRUE),
    health_conditions = paste(unique(na.omit(`field.plant.health.condition`)), collapse = ", "),
    .groups = "drop"
  ) %>%
  filter(n_health_conditions > 1) %>%
  arrange(`field.plant.number..from.tag.or.map.`)
kable(Health.Conflict.Summary[c(1:3,5)])

```

\newpage


## Observations where conditions got worse


```{r}
# Good <-> Fair

good_fair <- Health.Conflict.Summary %>%

  filter(health_conditions %in% c("Good, Fair", "Fair, Good"))

# Fair <-> Poor

fair_poor <- Health.Conflict.Summary %>%

  filter(health_conditions %in% c("Fair, Poor", "Poor, Fair"))

# Dead transitions

dead_transition <- Health.Conflict.Summary %>%

  filter(str_detect(health_conditions, "Dead")) %>%

  filter(

    str_detect(health_conditions, "Poor") |

    str_detect(health_conditions, "Fair")

  )

```


### Good <-> Fair
```{r}
kable(good_fair)
```


\newpage

### Fair <-> Poor

```{r}
kable(fair_poor)
```


\newpage

### Dead transitions

```{r}
kable(dead_transition)
```


\newpage

### All records with Dead

```{r}

All_Dead <- data %>%

  filter(`field.plant.health.condition` == "Dead") %>%

  group_by(`field.plant.number..from.tag.or.map.`, scientific_name) %>%

  summarise(

    n_dead_records = n(),

    n_total_records = n_distinct(observed_on), # replace with your date field if desired

    .groups = "drop"

  ) %>%

  arrange(`field.plant.number..from.tag.or.map.`)

kable(All_Dead)
```

\newpage



## Plants with insect activity

```{r}
insect.activity <- data %>% 
  filter(
    field.woody.tissue...insect.activity != "No Selection" & 
    field.woody.tissue...insect.activity != "None" &
    field.woody.tissue...insect.activity != "Not Sure" &
    field.woody.tissue...insect.activity != "" 
  ) %>%
  droplevels() %>%
  group_by(field.plant.number..from.tag.or.map.,scientific_name) %>%
  summarize(
    n_entries = n(),
    n_signs.of.disease = n_distinct(field.woody.tissue...insect.activity, na.rm = TRUE),
    insect.activity = paste(unique(na.omit(field.woody.tissue...insect.activity)), collapse = ", "),
    .groups = "drop"
    ) %>%
  arrange(field.plant.number..from.tag.or.map.)
kable(insect.activity[c(1,2,5)])
```


\newpage

## Plants with stem disease


```{r}
signs.of.disease <- data %>% 
  filter(
    field.woody.tissue...signs.of.disease != "No Selection" & 
    field.woody.tissue...signs.of.disease != "None" &
    field.woody.tissue...signs.of.disease != "Not Sure" &
    field.woody.tissue...signs.of.disease != "" 
  ) %>%
  droplevels() %>%
  group_by(field.plant.number..from.tag.or.map.,scientific_name) %>%
  summarize(
    n_entries = n(),
    n_signs.of.disease = n_distinct(field.woody.tissue...signs.of.disease, na.rm = TRUE),
    signs.of.disease = paste(unique(na.omit(field.woody.tissue...signs.of.disease)), collapse = ", "),
    .groups = "drop"
    ) %>%
  arrange(field.plant.number..from.tag.or.map.)
kable(signs.of.disease[c(1,2,5)])
```

